DENVar: A global Dynamic Estimate of NDVI Variance

### 2025-05-16 writing/manuscript.Rmd
added some refs and example fig for mean and var

---
title: "DENVar: A global Dynamic Estimate of NDVI Variance"
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'freshwater-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
```

@@

\noindent Since its early uses in the 1970s, the Normalized Difference Vegetation Index (NDVI) has greatly expanded our ability to quantify and monitor landscapes using satellite imagery

- NDVI is an estimate of greenness, NPP, forage availability
- NDVI has many uses and is often convenient because it's free, well-established, and

@@

@@

@@

* variance is hard to calculate, as it depends on a correct estimate of the mean
* paper on estimating the asymptote of environmental variance globally, but variance for a closed system must converge towards a finite value [@bachmaier_variogram_2011; @fleming_fine-scale_2014]

@@

```{r, eval=FALSE}
# approximate estimate of the number of unique DOY values
length(unique(lubridate::yday(as.Date('1981-09-01') + seq(0, 15800, by = 15))))
```

@@

@@

```{r hex-plots, fig.cap="Hex plots of the relationship between the global Dynamic Estimate of NDVI Variance (DENVar) and common ecological metricts. The fill of the hexagonal cells represents the log-10-transformed number of points within the cell. Rowwise, the variables are: mean NDVI (as estimated in this manuscript); the machine-learning Human Footprint Index (ml-HFI) of Keys, Barnes \\& Carter (2021); the 2015 cumulative, minimum, and seasonal range in the Dynamic Habitat Index of XXX, (F)."}
knitr::include_graphics('../figures/hexplots-canada-test.png')
```

@@

>>>>

---
title: "DENVar: A global Dynamic Estimate of NDVI Variance"
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Gavin L. Simpson
    institute: [au]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - au: Department of Animal & Veterinary Sciences, Aarhus University, Denmark.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'freshwater-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center', message = FALSE, warning = FALSE, comment = '#')
```

@@

\noindent Since its early uses in the 1970s, the Normalized Difference Vegetation Index (NDVI) has greatly expanded our ability to quantify and monitor landscapes using remotely sensed imagery [@pettorelli_using_2005; @pettorelli_normalized_2011; @anyamba_historical_2012]. NDVI is an estimate of a landscape's greenness, and it often a good proxy of primary productivity, forage availability, and vegetative cover [@pettorelli_normalized_2011; @merkle_large_2016]. However, as NDVI a proxy, it is not a direct measurement of such variables, and it should be interpreted carefully. NDVI is prone to loss of sensitivity in highly saturated (i.e., very green) environments [@matsushita_sensitivity_2007], and estimates can vary substantially across sensors [@huang_commentary_2021]. Still, NDVI is often used because it is free, well-established, and available at a wide variety of spatiotemporal scales [@ref].

@@

NDVI has been particularly useful when studying the movement behavior of herbivorous mammals, such as when testing the relationship between resource availability and range size [@ref]

NDVI has also been used to test the "green wave hypothesis", which was first developed by @ref in relation to migratory birds and states that animals will time their migration with the early green-up phases of forage growth to maximize forage nutrient content.

@@

scale in remote sensing and GIS: @goodchild_scale_2023

@@

* variance is hard to calculate, as it depends on a the estimated mean
* paper on estimating the asymptote of environmental variance globally, but variance for a closed system must converge towards a finite value [@bachmaier_variogram_2011; @fleming_fine-scale_2014]

```{r fig-smooth-wiggly, cache=TRUE, fig.cap="Examples of smooth and wiggly estimated trends in simulated NDVI data (points). Smooth estimates of the mean (top left) result in wiggly and large variance estimates (bottom left), as short-term trends in the data are assumed to be extreme events. In contrast, wiggly (i.e., flexible) estimates of the mean (top right) result in smooth and small variance estimates (bottom right), since short-term trends in the data are assumed to be trends in the mean. The data is the same in both columns.", fig.width=9, fig.height=4.5}
library('dplyr')   # for data wrangling
library('tidyr')   # for data wrangling
library('mgcv')    # for modeling
library('ggplot2') # for plotting
source('../analysis/figures/000-default-ggplot-theme.R')

# OU process given samples times, reversion time, standard deviation 
ou <- function(t, theta, sigma){
  n <- length(t) # number of samples
  dw  <- rnorm(n = n, mean = 0, sd = sqrt(sigma)) # noise between steps

  x <- numeric(n)
  x[1] <- 0 # set starting point
  for (i in 2:n) {
    x[i] <- (1 - theta) * x[i-1] + dw[i - 1]
  }
  return(x)
}

set.seed(3)
d <- tibble(t = seq(0, 1, length.out = 400),
            mu = (sinpi(sqrt(t + 0.5) * 10 - 0.5) + 1 * sqrt(t)) / 50 + 0.3,
            s2 = cospi(sqrt(t) - 0.5) * 0.001^2,
            noise = ou(t = t, theta = 0.2, sigma = sqrt(s2)),
            y = mu + noise)

if(FALSE) {
  ggplot(d) +
    geom_line(aes(t, mu), color = 'red') +
    geom_line(aes(t, y), alpha = 0.5)
}

m_wiggly <- gam(list(y ~ s(t, k = 30, bs = 'ad'), ~ s(t)),
                family = gaulss(), data = d, method = 'REML')
m_smooth <- gam(list(y ~ s(t, k = 10), ~ s(t)),
                family = gaulss(), data = d, method = 'REML')

get_preds <- function(mod, prefix = 'mod_') {
  predict(mod, newdata = tibble(t = seq(0, 1, by = 1e-3)),
          type = 'response', se.fit = FALSE) %>%
    as.data.frame() %>%
    transmute(Mean_est = V1, Variance_est = 1 / V2^2) %>%
    rename_with(.fn = \(.x) paste0(prefix, .x), .cols = 1:2)
}

bind_cols(tibble(t = seq(0, 1, by = 1e-3)),
          get_preds(m_wiggly, 'Wiggly_'), get_preds(m_smooth, 'Smooth_')) %>%
  pivot_longer(-t, names_sep = '_',
               names_to = c('model', 'lab', 'measure')) %>%
  pivot_wider(names_from = measure, values_from = value) %>%
  mutate(model = paste(model, 'estimate')) %>%
  ggplot() +
  geom_point(aes(t, y), tibble(t = 0, y = 0, lab = 'Variance'),
             color = 'transparent') +
  facet_grid(lab ~ model, scales = 'free', switch = 'y', ) +
  geom_point(aes(t, y), mutate(d, lab = 'Mean'), alpha = 0.1) +
  geom_line(aes(t, est), linewidth = 0.75) +
  ylab(NULL) +
  scale_x_continuous('Time', breaks = NULL) +
  theme(strip.placement = 'outside', strip.background.y = element_blank(),
        strip.text = element_text(size = 10),
        axis.title.x = element_text(size = 10))
```

@@

```{r n_doys, eval=FALSE}
# approximate estimate of the number of unique DOY values
length(unique(lubridate::yday(as.Date('1981-09-01') + seq(0, 15800, by = 15))))
```

@@

figures:

- plot grid of mean over space, `s(year)`, `s(doy)`, `s(elev)`
- plot grid of DENVar over space, `s(year)`, `s(doy)`, `s(elev)`, `s(mean)`
- histograms of DENVar around flat map of biomes (change colors)
- hex plots of DENVAr with other variables (add temperature, annual precip, extreme weather events, fires, invasive species, max body size, disease occurrence, ...)

@@

```{r hex-plots, fig.cap="Hex plots of the relationship between the global Dynamic Estimate of NDVI Variance (DENVar) and common ecological metricts. The fill of the hexagonal cells represents the log-10-transformed number of points within the cell. Rowwise, the variables are: mean NDVI (as estimated in this manuscript); the machine-learning Human Footprint Index (ml-HFI) of Keys, Barnes \\& Carter (2021); the 2015 cumulative, minimum, and seasonal range in the Dynamic Habitat Index of XXX, (F)."}
knitr::include_graphics('../figures/hexplots-global-test.png')
```

@@

supplementary material:

- hex plots of pred vs obs for mean and var
- histograms of mean around flat map of biomes (change colors)
